---
title: "Reading and Discussion"
subtitle: "Actionable items to modern scientific transparency"
format: stat159slides-revealjs
chalkboard: true
---

# Open Science and Reproducibility

## Reading

::: {.nonincremental}
- Title: **Seven Easy Steps to Open Science (An annotated reading list)**
- Authors: Sophia Cruwell et al.
- Journal: Zeitschrift fur Psychologie, 227(4), 237-248
- Year: 2019
- <https://doi.org/10.1027/2151-2604/a000387>
:::


# Introduction

## Why Are We Here?

- Current scientific incentive structures are heavily **misaligned** with transparency.

- Standard research practices frequently reward *novel, positive* results over rigorous, 
verifiable workflows.

- **The Solution:** Open science is a concrete set of behaviors that protect scientific 
credibility.


# Overview

## Step 1: Open Science Core Concepts

::: {.callout-note}
### The Core Concept

Open science is the process of making both the **content** and the **process** of 
producing scientific evidence fully transparent and accessible.
:::

- **Reproducibility Standard:** Original data + original code = identical original results.
- **Replicability Standard:** New data + identical experimental procedures = same findings.
- **Targeting Human Bias:** Acts as a procedural defense wall against confirmation bias.
- **Shifting Metrics:** Replaces shallow journal impact factors with actual pipeline audits.

::: {.notes}
🌐 Step 1: Open Science (Understanding the Concept)

* The Baseline Definition: It is the explicit process of making both the content and the entire process of producing scientific evidence transparent and accessible to others [INDEX].
* The Reproducibility Standard: You must guarantee that an independent researcher can take your exact data and code, run the same analysis, and obtain identical results [INDEX].
* The Replicability Standard: Your documentation must be clear enough that repeating your exact procedures with new participants yields the same findings [INDEX].
* Targeting Human Bias: It serves as a necessary procedural defense wall against human cognitive biases, such as confirmation bias and post hoc self-deception [INDEX].
* Shifting Career Metrics: Adopting these behaviors helps shift institutional hiring, funding, and promotion models away from shallow journal impact factors toward transparent metrics [INDEX].
:::


## Step 2: Open Access (OA)


::: {.columns}
::: {.column width="50%"}
### 🟡 The Gold Route

Works are made immediately and permanently free to the public by the publisher at the 
point of publication.
:::

::: {.column width="50%"}
### 🟢 The Green Route

Authors independently self-archive their peer-reviewed manuscripts or preprints in 
stable public repositories.
:::
:::

- **The Citation Boost:** OA articles observe an immediate tracking increase in 
citations between 36% and 600%.
- **Automated Audits:** Open plain text enables advanced data mining and field-wide 
meta-research.

::: {.notes}
🔓 Step 2: Open Access (Disseminating Research)

* The Main Objective: Removing all systemic and economic barriers to accessing, distributing, and re-using published scientific findings online [INDEX].
* The Gold Route: Publishing your work in an exclusively open-access or hybrid journal where it is made instantly free to the public by the publisher [INDEX].
* The Green Route: Self-archiving your work for free by uploading your peer-reviewed manuscripts or preprints to public, stable repositories (e.g., PsyArXiv) [INDEX].
* The Citation Advantage: Open-access articles receive a significant boost in professional utility, tracking between a 36% and 600% increase in citations [INDEX].
* Enabling Meta-Research: Exposing plain text openly allows data scientists to use automated text- and data-mining tools to audit field-wide claims [INDEX].
:::


## Step 3: Open Data, Materials, & Code

```{.yaml code-line-numbers="true" code-line-numbers-highlight="3-6"}
# The F.A.I.R. Ingestion Pipeline
data-management:
  Findable: Permanent DOI assignment (e.g., via OSF)
  Accessible: Unrestricted cloud repository locations
  Interoperable: Non-proprietary formats (use .csv, not .xlsx)
  Reusable: Comprehensive documentation and read-me file tracking
```

- **Data Minimum:** Moving away from "trusting" summaries to exposing raw datasets.
- **Materials Coverage:** Sharing every survey stimulus, software setup, and protocol to reconstruct the timeline.
- **Documentation Imperative:** Leaving a clean trail so your future self can understand column labels.
- **Anonymization:** Explicitly mapping exceptions when sensitive files cannot be shared.


::: {.notes}
📊 Step 3: Open Data, Materials, and Code (Transparency Fundamentals)

- The Minimal Baseline: Moving away from "trusting" a scientist's written summary to instead making the raw data files behind the analysis completely public [INDEX].
- Sharing Materials: Providing every document, experimental task, presentation tool, and digital stimulus required to reconstruct the exact study timeline [INDEX].
- The FAIR Data Protocol: Ensuring that your shared assets are structured to be Findable, Accessible, Interoperable, and Reusable over long periods [INDEX].
- The Documentation Imperative: Writing detailed read-me files so that outside users (and your own future self) can understand your columns and file paths [INDEX].
- Anonymization & Exceptions: Safeguarding participant privacy through careful data scrubbing, or explicitly reporting clear ethical boundaries when data cannot be shared [INDEX].
:::


## Step 4: Reproducible Analyses

- **Lab Technique Rule:** Computing best practices are as integral to robust research as 
fundamental laboratory methods.
- **The Raw Sanctuary:** Save your raw data completely separately from manipulated or 
cleaned iterations.
- **Structural Folders:** Maintain isolated directories per project, separating raw data, 
processed figures, and text scripts.
- **Record Everything:** Code every step of your analytical data adjustments in open-source 
scripts (R or Python).
- **Stable Archiving:** Deploy your finished code blocks to reputable, DOI-issuing hubs 
like Figshare, Dryad, or Zenodo.


::: {.notes}
💻 Step 4: Reproducible Analyses (Scientific Computing Skillsets)

* Computing as a Core Skill: Treating computing best practices as just as integral to high-quality science as fundamental laboratory techniques [INDEX].
* The Raw Data Sanctuary: Saving your raw, untouched data files completely separate from any cleaned, manipulated, or transformed versions [INDEX].
* Structural Project Folders: Dedicating a single isolated directory to each research project, with clear sub-directories for raw data, processed data, and text [INDEX].
* Recording Processing Steps: Tracking every single step of data cleaning and analysis by coding your pipeline in open-source languages (like R or Python) [INDEX].
* Stable DOI Archiving: Uploading your code scripts and research products to reputable, permanent, DOI-issuing repositories like Figshare, Dryad, or Zenodo [INDEX].
:::


## Step 5: Preregistration & Registered Reports

::: {.callout-warning}
### The Core Target Failure

Preregistration draws an unalterable boundary between planned **confirmatory** tests and unplanned **exploratory** data mining. It systematically blocks HARKing.
:::

- **Pre-Data Lock:** Committing to hypotheses and statistical plans *before* gathering or viewing data files.
- **Registered Reports:** Submitting your study design to a journal for peer-review **prior to data collection**.
- **In-Principle Acceptance:** The journal guarantees publication based on design rigor, completely regardless of the final outcome.

::: {.notes}
⏱️ Step 5: Preregistration and Registered Reports (Constraining Flexibility)

* The Pre-Data Mandate: Committing to your hypotheses, study designs, and statistical analysis plans before you gather data or view the files [INDEX].
* Demarcating Exploration: Forcing a transparent, unalterable boundary between planned confirmatory testing and unplanned, post hoc exploratory data mining [INDEX].
* The Target Failure: Restricting a researcher's degrees of freedom to eliminate HARKing (Hypothesizing After the Results are Known) and cherry-picking [INDEX].
* The Registered Report Pivot: Submitting your study design to a journal for pre-data peer-review to secure an "in-principle acceptance" before execution [INDEX].
* The Stage 2 Audit: Guaranteeing publication based purely on the rigor of the design and adherence to the plan, completely regardless of whether the final results are positive [INDEX].
:::


## Step 6: Replication Research

- **Benchmark of Truth:** An effect *must* be repeatable to count as a genuine scientific 
discovery.
- **Direct Replications:** Intentionally duplicating all critical design blocks to verify 
the physical result.
- **Conceptual Replications:** Purposefully altering operational variables to test 
the generalizability of a theory.
- **The "Alternative Test" Shift:** Re-labeling conceptual replications as 
*alternative tests* to stop muddy conflation.
- **Systemic Stabilization:** Adjusts career environments that unhealthily 
reward novelty over long-term stability.


::: {.notes}
🔬 Step 6: Replication Research (Verifying Scientific Discoveries)

* The Benchmark of Truth: Adopting the foundational scientific rule that an effect must be completely repeatable to count as a genuine scientific discovery [INDEX].
* Direct Replications: Intentionally duplicating every necessary element of an original study to see if the physical result can be reproduced [INDEX].
* Conceptual Replications: Purposefully changing operational components (like the measure or the sample) to evaluate the generalizability of the theory [INDEX].
* The "Alternative Test" Label: Re-labeling messy conceptual replications as "alternative tests" to stop them from being conflated with close, direct replications [INDEX].
* Systemic Stabilization: Embracing replication to fix structural market misalignments where fields reward novelty over long-term theory stability [INDEX].
:::


## Step 7: Teaching Open Science

- **The Educational Gap:** A vast majority of students currently graduate from 
statistics courses unaware of the replication crisis.
- **The 1-Hour Footprint:** Deploying compact, structured introductory modules 
radically alters how students view media claims.
- **The Behavioral Result:** Students learn to trust unverified research less, 
but view computational fields as closer to hard sciences.
- **Syllabus Integration:** Drastically cut instructor burden by leveraging 
pre-built infrastructure frameworks like FORRT or the Open Science MOOC.

::: {.notes}
🎓 Step 7: Teaching Open Science (Sustaining the Movement)

* The Educational Gap: Addressing the reality that a vast majority of students currently graduate from statistics courses having never heard of the replication crisis [INDEX].
* A 1-Hour Practical Footprint: Deploying structured 1-hour introductory modules to completely pivot how incoming researchers view media claims and data transparency [INDEX].
* The Dual Psychological Impact: Training students to trust unverified research less while simultaneously causing them to view their field as closer to the hard natural sciences [INDEX].
* Training Future Consumers: Realizing that open science pedagogy is vital even for students who leave research, creating critical public consumers of data [INDEX].
* Leveraging Existing Infrastructure: Utilizing free, collaborative open platforms like FORRT and the Open Science MOOC to drop pre-built resources straight into existing syllabi [INDEX].
:::


# Activity

## Set-Up

- Form groups of 3-4 students.
- Pretend your group is given $1,000 in "Academic Grant Money"
- I will show you a study context.


## Study Preregistration

You are going to conduct a study among developers tracking---during a week---their daily
[coffee consumption]{.mdlit} and [coding speed]{.hilit}.

_A few data rows shown below._

```{r}
set.seed(159)
n <- 100

days <- c("Mon", "Tue", "Wed", "Thu", "Fri")

dat = data.frame(
  developer = rep(LETTERS[1:20], each = n/20),
  day = rep(days, 20),
  coffee = runif(n, 0, 10),
  speed = runif(n, 30, 70)
)

head(dat, n = 10)
```

## Study Preregistration

Write down your Null and Alternative hypotheses.
In other words, write down what kind of relationship you would expect to see in 
your study.

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

## Exploratory Analysis (part I)

## Exploratory Analysis (part I)

```{r}
library(ggplot2)

ggplot(dat, aes(coffee, speed)) +
  geom_point() +
  theme_bw() +
  labs(title = "Consumption Mon-Fri",
       subtitle = "Correlation 0.05", 
       x = "Coffee consumption",
       y = "Coding Speed")
```

::: {.notes}
Scatter plot showing that coffee consumption actually had zero effect on coding speed.
:::


## Exploratory Analysis (part II)

## Exploratory Analysis (part II)

```{r}
set.seed(159)
m <- 20
x <- runif(m, 0, 10)
y <- 10 * sqrt(x + rnorm(m, mean = 5, sd = 3))
tue = data.frame(coffee = x, speed = y)

ggplot(tue, aes(x = coffee, y = speed)) +
  geom_point() +
  theme_bw() +
  labs(title = "Consumption on Tuesday",
       subtitle = "Correlation 0.65", 
       x = "Coffee consumption",
       y = "Coding Speed")
```

## Exploratory Analysis (part II)

\

Developers who drank espresso specifically on Tuesdays coded 40% faster.


## Auction

- The prestigious [_Journal of Caffeinated Studies_]{.bold-hilit} is accepting submissions.
- But they __only__ publish statistically significant findings.
- You must "bid" your grant money to publish.

. . .

\

:::: {.columns}
::: {.column}
Stick to your original hypotheses
:::

::: {.column}
Decide to HARK on the "Tuesday Espresso" anomaly
:::
::::


::: {.notes}
HARK: Hypothesizing After the Results are Known
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```


## BTW

The "Tuesday Espresso" spike is actually just a noise artifact. 

If you spent your money to publish it, you just committed a massive, 
irreproducible false-positive sin 🤭

::: { .notes}
Students experience the intense system pressure to choose flashy data-mining over 
pre-planned truth, instantly contextualizing why Step 5 Preregistration is so vital.
:::


# Poll & Quiz

## Poll

Rate your level of agreement with this statement: 

A junior researcher has a moral obligation to release their raw data under a CC0 
(Public Domain) license, even if an outside lab can use it to scoop their next publication.

::: {.nonincremental}
1) Strongly Disagree (Code or nothing)
2) Disagree
3) Neutral
4) Agree
5) Strongly Agree (Spreadsheets are essential)
:::


## Poll

Where does the majority of true, groundbreaking scientific progress actually happen?

::: {.nonincremental}
A) Purely Confirmatory: Following a rigid, preregistered blueprint flawlessly to eliminate 
human bias.

B) Purely Exploratory: Wandering through the data landscape, discovering unexpected 
patterns by accident.

C) Mostly Confirmatory, but keeping a hidden "sandbox" for post hoc visual data exploration.

D) They are completely equal, and separating them into strict categories hurts the 
scientific mind.
:::

::: {.notes}
The question in this slide involves unpacking the philosophical tension between 
strict confirmatory scripts and raw exploratory intuition
:::


## Poll

Imagine you are in charge of a federal funding agency that has to allocate a limited 
tax-payer budget.

You have $10,000,000 in federal grant money to stabilize an irreproducible field.
How do you split your funding?

::: {.nonincremental}
A) [100% to Direct Replications:]{.bold-hilit} We must verify if the original computer 
code and methods are rock-solid first.

B) [100% to Alternative Tests (Conceptual):]{.bold-hilit} We must verify if the underlying 
scientific theory actually applies to different real-world scenarios.

C) [50/50 Split:]{.bold-hilit} Balance code verification with conceptual generalizability.

D) [0% to Replication:]{.bold-hilit} Spend it all on funding completely brand-new, 
novel discoveries instead.
:::


## Poll

If you were forced to align your upcoming research career with one of these two profiles 
right now, which one describes you?

:::: {.columns}
::: {.column}
[Career Pragmatist]{.bold-hilit}

My priority is survival, secure funding, and graduation. 
I will adopt open-science tools only if they don't slow down my publishing 
timeline or put my project at risk of being scooped
:::

::: {.column}
[Open-Science Idealist]{.bold-hilit}

My priority is absolute scientific truth and provenance. I would rather delay a 
publication, or release my raw data under a CC0 license if it means my pipeline is 
perfectly transparent
:::
::::

\

If you had to choose one of the positions, which one would you choose?


# Chafa Stuff

## Shifting from Trust to Transparency

INSERT IMAGE OF TRUST FALL


## The Ultimate Scientific Standard

* Slide Title: Open Science: Redefining How We Know What Is True
* Visual Anchor: A split screen showing a newspaper headline of a flashy 
"scientific discovery" on one side, and a complex flowchart of code/data on the other.
* Core Concepts:
* The Baseline: Replication is the ultimate standard of scientific truth (Seven_easy... p. 3).
   * The Problem: In complex, data-heavy modern science, full replication is often incredibly expensive, time-consuming, or physically impossible (Seven_easy... p. 3).
   * The Solution: Open Science bridges the gap by making the internal process visible (Seven_easy... pp. 1, 3).

::: {.notes}
Today we are talking about a massive shift happening right now in research. 
Historically, science operated on trust—we read a paragraph in a journal and 
assumed the authors did the math right. Today, we are moving to a model of 
verifiability (Seven_easy... p. 4). If you can't show me the data and the steps, 
it didn't happen (Seven_easy... pp. 3, 5)."
:::


## The Credibility Revolution & Its Threats

* Visual Anchor: A simplified reproduction of Figure 1 (Threats to reproducible 
science) from the paper (Seven_easy... p. 3).
* Core Concepts:
* Publication Bias: Journals reward clean, positive headlines; null/negative results get locked away in the "file drawer" (Seven_easy... p. 3).
   * P-Hacking & HARKing: Subconsciously or consciously manipulating statistical models or changing hypotheses after looking at the data to fit a narrative (Seven_easy... p. 3).
   * The Crisis Metric: Estimates suggest a staggering amount of published literature in fields like psychology and biology struggles to replicate (Seven_easy... p. 1).

::: {.notes}
Why are we doing this? Because human brains are naturally wired for 
self-deception (Seven_easy... pp. 3, 6). Under the pressure of a deadline 
or a grade, it is terrifyingly easy to tweak a filter or drop a few outliers 
to get a 'significant' result (Seven_easy... p. 3). We aren't fighting bad 
intentions; we are fighting flawed workflows (Seven_easy... p. 3).
:::


## Open Science is a Set of Behaviors

* Step 1: Open Access (Green vs. Gold) (Seven_easy... pp. 3-4)
* Step 2: Open Data, Materials, & Code (Seven_easy... pp. 1, 4)
* Step 3: Reproducible Computing Workflows (Seven_easy... pp. 1, 5)
* Step 4: Preregistration & Registered Reports (Seven_easy... pp. 1, 6-7)
* Step 5: Replication Tracking (Seven_easy... pp. 1, 7)

* Open Science is not an "all-or-nothing" bureaucratic pass/fail checklist 
* It is a fluid spectrum of behaviors that you build into your daily research 
stack (Seven_easy... pp. 1-3).

::: {.notes}
The paper breaks this monster concept down into modular, actionable steps 
(Seven_easy... pp. 1-2). You don't have to change your entire lifestyle overnight. 
You just need to change your habits. It starts with how you organize your desktop 
folders and how you name your files (Seven_easy... p. 5)."
:::


## The Selfish Return on Investment (ROI)

* Visual Anchor: A graphic balancing an "Academic Reputation Shield" against a 
"Time Tracker" showing hours saved.

* Computational Insurance: Flawless file management keeps raw data safe from 
corruptions that cause high-profile retractions (Seven_easy... p. 5).
* The Citation Magnet: Openly sharing code and data means your work is 
trusted and cited significantly more by peers (Seven_easy... pp. 4-5).
* Future-Self Empathy: Writing "executable descriptions" (clean scripts) means 
you can rerun your own work six months from now without crying (Seven_easy... p. 5).

::: {.notes}
Forget altruism for a second. Let's talk about why this is great for your own 
sanity (Seven_easy... p. 5). When you automate your pipeline, you create an 
unalterable paper trail (Seven_easy... pp. 5-6). If a peer finds an anomaly 
in your work, an automated script proves it was an innocent typo, saving your 
career (Seven_easy... p. 5). If you used a manual Excel copy-paste method, 
you look like a fraud (Seven_easy... p. 5).
:::


## The Challenge & Handoff

* Transitioning from Consumers to Auditors
* Visual Anchor: A bold text callout reading: "Data is Cheap. Time is Expensive." 
(Seven_easy... pp. 5-6)
* The goal of this lecture block is to learn how to actively audit research, 
not just passively read it (Seven_easy... pp. 1, 8).
* Setting up the rules of the upcoming classroom debate and scenario challenges.

::: {.notes}
As we move into our discussion, I want you to stop acting like passive consumers 
of data (Seven_easy... p. 8). You are now the gatekeepers. We are going to 
stress-test these seven steps against real-world scenarios to see where workflows 
shatter, and how we can engineer a more transparent future (Seven_easy... pp. 1-2). 
Let's jump in.
:::


# Reflection Points

## Reflections

- Step 1 (Open Science Overview): Is open science a rigid set of bureaucratic 
rules you must pass, or is it a fluid collection of daily behaviors you choose 
to practice (Seven_easy... p. 2)?

\

- Step 2 (Open Access): If a university library pays millions for journal access 
while low-income global scholars are locked out, who actually owns the knowledge 
produced by public funds (Seven_easy... p. 4)?

\

- Step 3 (Open Data & Code): If you claim your data is available 'upon reasonable 
request' but you ignore peer emails five years later, did you actually share your 
data (Seven_easy... p. 4)?


## Reflections

- Step 4 (Reproducible Analyses): If a peer uses your exact dataset but has to 
manually guess how you clicked your way through an SPSS or Excel menu, is your 
science truly transparent (Seven_easy... pp. 5-6)?

\

- Step 5 (Preregistration): Why do human brains naturally look at accidental 
patterns in random noise and confidently declare: 'I knew that would happen all 
along' (Seven_easy... p. 6)?

\

- Step 6 (Replication Research): Why does the modern academic system celebrate 
an unverified, flashy headline over a quiet, rock-solid verification study 
(Seven_easy... pp. 3, 7)?

