---
title: "Dare to Share"
subtitle: "The Code-Sharing Paradox"
format: stat159slides-revealjs
chalkboard: true
---

# Share Your Code

## Reading 1

::: {.nonincremental}
- Title: **Top ten reasons to not share your code (and why you should anyway)**
- Authors: Randall J. LeVeque.
- Journal: Software Sustainability Institute
- Year: 2013
- <https://www.software.ac.uk/blog/top-ten-reasons-not-share-your-code-and-why-you-should-anyway>
:::


## Reading 2

::: {.nonincremental}
- Title: **Publish your computer code: it is good enough**
- Authors: Nick Barnes.
- Journal: Nature 467 (7317): 753 
- Year: Oct 2010
- doi: 10.1038/467753a
- <https://www.software.ac.uk/blog/top-ten-reasons-not-share-your-code-and-why-you-should-anyway>
:::


# Introduction

## Why Are We Here?

- [The Reality:]{.bold-hilit} Most computational research papers publish beautifully polished figures 
but hide the underlying source scripts.

\

- [The Excuses:]{.bold-hilit} *"My code is too messy,"* *"Competitors will steal it,"* or 
*"I don't have time to clean it up."*

\

- [Today's Goal:]{.bold-hilit} We are going to stress-test these justifications using the arguments of 
Nick Barnes (2010) and Randall LeVeque (2013).


# Overview: Barnes' Paper

## Most Software Isn't Very Good

- [A Professional Perspective:]{.bold-hilit} Nick Barnes, a professional software engineer, 
shares a trade secret: most professional software inside consumer electronics is badly 
documented, inconsistent, and poorly tested.

- [The Scientist's Assumption:]{.bold-hilit} Scientists generally withhold their code 
because they believe it is poor—lacking good comments, sensible variable names, 
or proper indentation.

- [The Reality Check:]{.bold-hilit} These formatting flaws **do not matter** to the 
core scientific record. Software across all trades is simply written to be 
*good enough for the job intended*.


## What Happens to Science Software?

- [The Computational Reality:]{.bold-hilit} Turning raw laboratory data into published 
research papers inevitably requires programming and data manipulation.

- [The Patchwork Pipeline:]{.bold-hilit} Science code is often an awkward mix of small 
scripts to draw charts, larger filtering routines, and telemetry software glued together 
with piecemeal scripts.

- [The Disappearing History:]{.bold-hilit} While papers include a brief mathematical 
description of an algorithm, the actual software code usually languishes in an obscure 
folder or is simply deleted after publication.


## Clean Enough vs. "Good Enough"

::: {.small}
```{.python code-line-numbers="true" code-line-numbers-highlight="3-4"}
# The Real-World Research Script
import pandas as pd
df = pd.read_csv("temp_data_FIXED_FINAL_v2.csv") # Ugly hardcoded name!
res = [x * 4.219 for x in df['score'] if x > 0] # Messy magic number hack!
print("It works, leave it alone.")
```
:::

- [The Psychological Barrier:]{.bold-hilit} Scientists frequently withhold code because 
they are embarrassed by non-standard syntax, messy comments, or unoptimized loops.

- [Ugly code is good enough:]{.bold-hilit} A dirty, functional script that exposes your 
actual workflow is infinitely superior to a hidden, "pristine" secret.

- [Not Commercial Software:]{.bold-hilit} Your code is an archive of an experiment, 
not an enterprise app. Readers do not expect commercial engineering standards—they 
expect scientific truth.


## Openness Improves the Science

- [Case Study: NASA & Climate Code]{.bold-hilit} In 2007, NASA released its global 
temperature reporting software online. The code was messy and difficult for critics 
to run locally.

- [The Volunteer Audit:]{.bold-hilit} Volunteers rewrote the messy software to make 
it easier for non-experts to run. Through this open inspection, they discovered minor bugs.

- [The Ultimate Outcome:]{.bold-hilit} The bugs had zero bearing on the final scientific 
results, but NASA fixed them and ultimately chose to replace their original code with 
the volunteer version.

- [The Lesson:]{.bold-hilit} Science source code that is hidden from public scrutiny 
cannot benefit from peer review and rapid community improvements.


## "No More Excuses" for Secrecy

- ["It is not common practice."]{.bold-hilit} <br>
This is a cultural lag that must change across all fields.

- ["People will pick holes and demand support."]{.bold-hilit} <br>
Not publishing code draws allegations of fraud, which is far worse. Furthermore, 
nobody is entitled to technical support for freely provided code; you can simply ignore unhelpful feedback.

- ["The code is valuable Intellectual Property."]{.bold-hilit} <br>
Frankly, a little Python/R/Matlab routine is rarely worth money. Without ongoing backing by 
skilled experts, unshared institutional code just becomes un-useable "abandonware".


## Overcoming Computational Perfectionism

- To scientists, the word "publication" signifies perfectionism. 

- This psychological wall stops researchers from releasing functional code 
because they feel it isn't "polished" enough.

- When publishing a paper, the original, raw script can simply be published as 
supplementary online information.

- Ask yourself: _Does your code perform the algorithm you describe in your paper?_
If it does, your audience will accept it "as is". If it does not, you need to fix it anyway.


# Overview: LeVeque's Paper

## The "Alternative Universe" Analogy

LeVeque asks us to imagine a world where mathematicians publish brilliant theorems but keep 
their proofs completely secret, expecting readers to simply "trust their reputation."

In modern science, publishing a theorem without a proof is considered a joke. 

Yet, publishing a computational figure without its source script is standard operating procedure.


## The Hidden Value of Imperfect Scripts

- [Research Code vs. Enterprise Software]{.bold-hilit} <br>
Unlike commercial software, research code is not expected to be perfectly polished, 
but it is an essential part of the scientific record.

- Code contains crucial details not present in the paper, serving as necessary 
documentation for the study.

- Cleaning up code for public release is an effective way for researchers to uncover 
bugs in their own work.


## What If You Cannot Share?

- Partial Sharing: Even with proprietary restrictions, researchers can often share 
the specific, novel routines they developed for the paper.

- Inspection Over Execution: Simply being able to read and inspect code is often 
more valuable to reviewers than running a compiled, black-box executable.

- Alternative Mediums: If code cannot be shared digitally, publishing it as a static 
PDF appendix is a viable alternative.


## The Self-Correction Effect

Almost every researcher who forces themselves to clean up and document their code for a 
public release uncovers hidden bugs, memory leaks, or flawed assumptions before publication.

- [The Blind Review Blindspot:]{.bold-hilit} Peer reviewers cannot catch internal logic 
flaws or scrambled data rows simply by staring at a static PDF chart or a final table 
of p-values.

- [Open Inspection over Execution:]{.bold-hilit} For many peer reviewers, simply being able 
to **read** the logic of the source text is vastly more informative than trying to run a 
compiled file.

- [The Crowdsourced Audit:]{.bold-hilit} Making your script open to the world allows outside 
domain experts to spot anomalies early, preventing high-profile retractions later on.


## Transparency Maximizes Professional Impact

- [The Academic Currency:]{.bold-hilit} Code sharing is a major engine for professional 
discovery. Other scientists can only cite your work if they can easily adapt and extend 
your workflow.

- [Establishing Provenance:]{.bold-hilit} Publicly archiving your scripts on a hosted 
remote with a permanent DOI establishes clear, timestamped ownership over your unique 
algorithms.

- [Stopping Code Theft:]{.bold-hilit} If code sharing becomes standard operating procedure, 
it is significantly harder for competitors to secretly use your ideas without giving 
explicit citation credit.

- [Future Proof Competency:]{.bold-hilit} As the scientific community rapidly evolves, 
researchers who provide fully auditable pipelines are heavily favored by modern funding 
agencies and hiring committees.


# Discussion Time

## Activity

We're going to divide the classroom into groups

. . .

\

:::: {.columns}
::: {.column}
[Team A: The Proprietary Tech Firm]{.mdlit}

You have built a simulation code that calculates the structural stress limits for 
public bridges or medical equipment. You claim it's a corporate secret and refuse 
to share the code lines.
:::

::: {.column .fragment}
[Team B: The Public Safety Board]{.hilit}

You refuse to approve the engineering projects because you cannot verify the 
black-box simulation.
:::
::::


## Activity (cont'd)

- A bridge design built using the firm's secret code has just shown minor structural cracks. 
- The firm claims it's an isolated steel issue, not a software issue.
- They refuse to release their code lines, calling it a multi-million dollar corporate secret.

. . .

Huddle with your immediate neighbors to formulate 3 compelling arguments to defend 
your position.

```{r}
#| echo: false
countdown::countdown(5, bottom = 0)
```

<!--
The Proprietary Tech Firm

1. **[Argument 1: Intellectual Property]**
2. **[Argument 2: Support Overload]**
3. **[Argument 3: Code Modification Risks]**

- Commercial Survival: How do you protect your multi-year financial investment from 
being copied by global corporate rivals]?
- The User Support Trap: If the code goes public, how do you prevent your engineering 
team from being hijacked by strangers demanding free tech support]?
- The Liability Shield: If you open-source the code and someone modifies a parameter, 
causing a crash, who legally owns that failure?


The Public Safety Board:

1. **[Argument 1: Public Accountability]**
2. **[Argument 2: Forensic Transparency]**
3. **[Argument 3: Unverifiable Black Boxes]**

- The Black Box Danger: Why is an un-auditable computational simulation a fundamental 
violation of the scientific method?
- The Life-or-Death Reality: As LeVeque warns, when math moves "outside the ivory tower," 
what happens when code bugs carry real-world body counts?
- The Peer Review Minimum: Why is staring at a final PDF chart or commercial certificate 
completely empty without line-by-line inspection?
-->


## Debate

- Imagine you are a junior researcher. 
- Your script is unoptimized, un-commented, and full of messy hacks.
- But it works: it successfully generates your paper's plots. 
  
. . .

If you publish this "dirty" code:

:::: {.columns}
::: {.column width="45%"}
::: {.poll}
Do you elevate scientific transparency? 
:::
:::

::: {.column width="10%"}
\

&nbsp;&nbsp; OR
:::

::: {.column width="45%"}
::: {.poll}
Do you pollute the scientific record with amateur software engineering? 
:::
:::
::::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

::: {.notes}
If someone emails you for your code 3 years from now, why is that infinitely more 
painful than just dropping your dirty script onto GitHub today?"
:::


## Debate

. . .

Going back to LeVeque's "Theorems Without Proofs" analogy, publication of theorems without
proofs is not acceptable. Yet, we accept computational outputs without the code that 
produce them.

. . .

::: {.poll}
Why does the scientific community tolerate this double standard?
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

::: {.notes}
Is a computer script fundamentally different from a mathematical proof, 
or are we just living in a scientifically immature era of computing?
:::


## Debate

<!-- Protect Your Output vs. Public Provenance -->

::: {.columns}
::: {.column width="50%"}
### 🛡️ The Pragmatist View

It took me three years to write this algorithmic base. If I open-source it now, my 
competitors will use my code to publish five papers before I can even finish my next draft.
:::

::: {.column width="50%"}
### 🌐 The Open View

Publicly archiving your code with a timestamped DOI is the absolute best way to 
claim ownership, establish provenance, and legally force competitors to cite you.
:::
:::

. . .

::: {.poll}
If you are a funding agency, how do you balance a scientist's valid right to profit 
from their multi-year software labor against the public's right to verify the data 
pipeline immediately?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```


## Debate

<!-- The Fear of Maintenance -->

Many researchers hide their scripts because they are terrified that once the code is public, 
their inbox will be flooded with emails from strangers asking for technical troubleshooting, 
custom modifications, or software installation help.

. . .

::: {.poll}
How do you draw a boundary between providing necessary scientific transparency and protecting 
your personal time from being hijacked by outside users demanding free tech support?
:::

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```

::: {.notes}
Barnes and LeVeque emphasize that research code is a static archive of a publication, 
not a commercial product that comes with customer support. 
:::


# Quiz-Poll Time

## Google Form

\

**INSERT QR code!!!**

\

**INSERT <url> link to google form!!!**


## Poll

<!--The "Ugly Code" Reality Check -->

You are about to submit a paper. Your code works but is messy, uncommented, and unoptimized. What is your honest action?

::: {.nonincremental}
A) [Publish it immediately:]{.bold-hilit} The dirty code is an essential part of the 
scientific record; transparency wins over aesthetics.

B) [Hide it entirely:]{.bold-hilit} I am too self-conscious about my coding skills; 
I would rather hide the script than face professional judgment.

C) [Delay the paper:]{.bold-hilit} Spend three weeks rewriting it neatly, even if it delays 
my graduation or submission timeline.

D) [The "Upon Request" Lie:]{.bold-hilit} State that the code is "available upon request" 
and hope no one ever actually emails me.
:::


## Poll

<!-- The Computational Double Standard -->

Why does the scientific community tolerate papers that publish computational figures but hide the source code scripts?

::: {.nonincremental}
A) [Cultural Lag:]{.bold-hilit} Computing is relatively new to academia; we are simply living in a scientifically immature era of data science.

B) [Complexity:]{.bold-hilit} A computer script is infinitely more complex than a static math proof, making routine auditing impractical.

C) [Status Bias:]{.bold-hilit} We evaluate science based on the prestige and reputation of the author's university rather than their code.

D) [Journal Laziness:]{.bold-hilit} Editors and peer reviewers simply refuse to force authors to change their habits because it adds extra work.
:::


## Poll

<!-- The "Scooping" Threat Matrix -->

Rate your level of agreement with this statement: 

"A researcher has a right to keep their novel code hidden for up to two years after publication to ensure competitors don't scoop their future research ideas."


::: {.nonincremental}
1) Strongly Disagree (Public tax dollars funded it; code belongs to the world immediately)
2) Disagree
3) Neutral / Undecided
4) Agree
5) Strongly Agree (Career survival metrics must outweigh open-science ideals)
:::


## Poll

What is your biggest psychological fear regarding open-sourcing your personal computational 
research project?

\

In exactly **one word**, submit your primary worry or mental roadblock regarding making 
your code visible to the public.


# Final Words



## Key Takeaways

- Open science isn't just an academic debate—it has high-stakes civic consequences.

- Hidden code can carry direct life-or-death consequences in engineering and public policy.

- Perfectionism and Absolute Secrecy both break scientific progress. 


## Key Takeaways

In the real world, engineering projects navigate this standoff through compromised 
structures:

- **The Production Sandbox:** Tech firms can deposit source code into a 
    *Software Escrow*—allowing a trusted, independent third-party academic panel to 
    run line-by-line audits without exposing trade secrets to public competitors.

- **The Barnes Reality Check:** Code does not need to be a beautifully polished 
    product to be verified; it just needs to expose the raw mathematical parameters 
    governing the bridge's stress metrics.

- **The Baseline Rule:** Un-auditable code is un-verifiable science.


## The Selfish Benefits of Open Code

- [Your Most Important Collaborator:]{.bold-hilit} Your "past self" is your most frequent 
collaborator, and they notoriously forget what their own code did after 6 months.

- [The "Push-Button" Advantage:]{.bold-hilit} Forcing yourself to build a script that 
replicates your whole project with the push of a button pays off massively down the road.

- [Frictionless Revisions:]{.bold-hilit} When a journal reviewer inevitably demands you 
replot your data or shift an axis, an automated script saves you weeks of manual 
reconstruction labor.

- [The Institutional Shield:]{.bold-hilit} If your results are ever publicly questioned 
or audited by outsiders, a public, timestamped git repository provides immediate, 
undeniable proof of your data's provenance.
