---
title: "Data Science at the Singularity"
subtitle: "The Singularity Agenda"
format: stat159slides-revealjs
chalkboard: true
---

# Share Your Code

## Reading

::: {.nonincremental}
- Title: **Data Science at the Singularity**
- Authors: David Donoho
- Journal: Harvard Data Science Review, 6(1)
- Year: 2024
- DOI: https://doi.org/10.1162/99608f92.b91339ef
- <https://hdsr.mitpress.mit.edu/pub/g9mau4m0/release/2>
:::


# Introduction

## Why Are We Here?

- [The Shift:]{.bold-hilit} Computation-based research is transitioning toward 
**Frictionless Reproducibility (FR)**. 

- [The Triad:]{.bold-hilit} FR driven by maturity of 3 data science principles: 
    + data sharing
    + code sharing
    + competitive challenges

- [The Superpower:]{.bold-hilit} Empirical Machine Learning (EML) uses this triad as an 
emergent superpower to drive AI progress.

- [The Export:]{.bold-hilit} These practices can be learned and adopted by *any* research 
field to instantly accelerate its rate of progress.


# Overview

## The Old Way vs. The New Way

The Old Way (In-Principle Sharing)

- A scientist publishes a paper with a beautiful chart. 
- If you want to check it, you have to email them, 
- Wait weeks for the data, 
- figure out their messy code, 
- and fix a dozen errors. 

## The Old Way vs. The New Way

The New Way (Frictionless Reproducibility)

- You read a paper, 
- click a link, 
- clone their GitHub repository, 
- and run their exact analysis on your laptop in 2 minutes. 

. . .


The "Singularity" 

When checking other people's data work becomes effortless, 
scientific progress explodes at breakneck speed.


## Cheap and Massive Compute Power

- [The Stalled Engine:]{.bold-hilit} Around 2005, individual computer chips stopped getting 
significantly faster year-over-year.

- [The Smartphone Boom:]{.bold-hilit} Because billions of people bought smartphones, 
tech companies built a massive global network of cloud servers to handle all that data.

- [The Result for Data Science:]{.bold-hilit} Today, any undergraduate student with a laptop can rent 
a supercomputer on the cloud for pennies to run massive datasets. 
Space and processing power are no longer bottlenecks.


## Data Science Matures

The Core Triad:

1. [Open Data:]{.bold-hilit} Exposing the raw numbers, not just summaries.
2. [Open Code:]{.bold-hilit} Sharing the exact scripts so others can re-run the math.
3. [Common Challenges:]{.bold-hilit} A shared goal where everyone uses the same dataset to compete for the best accuracy scores on a public leaderboard.

. . .

**Phase Transition:** When all three legs of the triad meet, researchers can reproduce 
or modify an outside institution's results within minutes or clicks.


## What Happens If You Hide One Leg?

:::: {.columns}
::: {.column width="33%"}
### 🛑 No Challenges

- You get "Open Science" but no common goal. 
- Everyone works in isolated bubbles.
- Impossible to rank which method actually works best.
:::

::: {.column width="33%"}
### 🛑 No Code

- You get competitions with secret methods (like the Netflix Prize). 
- People win, but no one learns *how* they won because the code is locked.
:::

::: {.column width="33%"}
### 🛑 No Data

- You get "Bring-Your-Own-Data" science (like private medical trials). 
- You cannot verify the claims because the underlying data is a black box.
:::
::::


## Viral Science: Zero Friction

[The Meme Analogy]{.bold-hilit} <br>
Why do internet memes spread across the globe in a single afternoon? 
Because there is **zero friction** to share them—you just click "forward." 

- [Viral Code:]{.bold-hilit} When data, code, and leaderboards are completely open, 
scientific ideas spread through a research community exactly like viral memes.

- [The "URL or I Don't Care" Rule:]{.bold-hilit} Modern data analysts reject paywalls 
and broken links. If a research method doesn't have an immediate, working link, 
the community ignores it and moves on.


## Automated Science Tracking

- [Code as Gold:]{.bold-hilit} In the future, public code repositories will be treated as raw material. 

- [Automated Tweaking:]{.bold-hilit} Data scientists will write automated scripts designed 
to crawl through thousands of *past* papers, pull down their code, and automatically 
swap parameters to find hidden improvements.

- [The Backwater Risk:]{.bold-hilit} Research fields that refuse to share their data and 
code due to privacy or copyright fears will be left behind and forgotten by the next generation.



# Discussion Time


## Debate

<!-- Scientific Truth vs. Internet Memes -->

::: {.notes}
The paper argues that when you combine Open Data, Open Code, and Public Leaderboards, 
scientific ideas spread through a research field exactly like viral internet memes
that are instantly copied, tweaked, and reshared.
:::

Internet memes spread fast because they are funny, short, and effortless to share, 
not because they are "true." 
  
\

If we turn scientific research into a fast-paced, viral marketplace where the 
"easiest to download" code wins the day, do we risk prioritizing convenience 
over deep scientific accuracy? Does "viral science" breed shallow thinking?

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```


## Debate

<!-- Leaderboard Chasing vs. Deep Theory -->

::: {.notes}
The text mentions that mathematical purists often look at leaderboard-driven s
cience with "shock and disbelief". They argue that climbing a scoreboard isn't the 
same as understanding the underlying laws of nature.
:::

Imagine your lab builds an algorithm that gets a 99% accuracy score on a public dataset, 
winning 1st place on the leaderboard---but you cannot explain mathematically *why* it works. 

On the other side of the room, a classmate has a beautiful mathematical formula that explains 
the universe, but only hits 75% accuracy on the data.
  
**Who did better science? If you are a funding agency, who gets the million-dollar grant?**

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```


## Debate

<!-- Banning the "Black Boxes" -->

::: {.columns}
::: {.column width="50%"}
### 🔒 The Secret Data Problem

A research team has a revolutionary breakthrough in cancer detection, 
but the data belongs to a private hospital corporation and 
legally cannot be shared publicly.
:::

::: {.column width="50%"}
### 🌐 The Open Science Rule

Donoho notes that next-generation researchers operate on the rule "URL or I don't care." 
If they can't download the script and data instantly, they ignore it.
:::
:::

\

Should scientific journals completely ban papers that use proprietary or private "black box" data? 

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```


## Debate

<!-- Computational Amnesia -->

::: {.notes}
Donoho predicts that because the younger generation values immediate 
computational "flow," older scientific papers that lack working code repos 
or downloadable data links will be completely forgotten.
:::

::: {.nonincremental}
- Imagine a brilliant scientific breakthrough written in 1985. 
- It is full of incredible insights, but the math was done on paper, 
- And there is no GitHub link or CSV file to download.
:::

If our generation completely stops reading older, non-digital research because 
it has too much "computational friction", are we actually moving science forward, 
or are we intentionally blinding ourselves to the past?

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```


## Debate

<!-- Will Code Automation Kill Creativity? -->

::: {.notes}
The paper predicts that future data scientists won't spend time manually 
writing code line-by-line. Instead, automated AI systems will crawl public repositories, 
download past papers, and robotically swap parameters to find data improvements.
:::

If computers automate the tactical execution of science:

- running the scripts, 
- tweaking the code, 
- and optimizing the charts automatically

. . .

What is left for human scientists to do? 
  


# Quiz-Poll Time

## Google Form

\

**INSERT QR code!!!**

\

**INSERT <url> link to google form!!!**


## Poll


## Poll

<!-- The Meme Analogy Reality Check -->

If a research method becomes highly popular and "viral" simply because the code repository 
is incredibly easy to download and run, what is the biggest danger?

::: {.nonincremental}
A) [Shallow Science:]{.bold-hilit} Researchers will choose code that is convenient rather than 
code that is mathematically accurate.

B) [No Danger:]{.bold-hilit} Popularity is proof that it works; if a script is easy to use, 
it naturally helps more people do science.

C) [Hyper-Overfitting:]{.bold-hilit} Everyone will end up using the exact same three scripts, 
freezing all creative variation in the field.

D) [The Information Flood:]{.bold-hilit} It will generate too many low-quality, 
auto-generated papers for peer reviewers to handle.
:::


## Poll

<!-- The Leaderboard Standoff -->

You are a federal funding agency with a $1,000,000 grant. Two research teams submit proposals 
to fix an active environmental crisis. Who gets the money?

::: {.nonincremental}
A) [The Scoreboard Winner:]{.bold-hilit} The team with an empirical script hitting 99% predictive accuracy 
on a public test dataset—even though they cannot mathematically explain *why* it works.

B) [The Theoretical Master:]{.bold-hilit} The team with a beautiful, mathematically derived 
formula that perfectly explains the physics of the crisis, 
but only hits 75% predictive accuracy on the data.

C) [Split the Check:]{.bold-hilit}* Give them $500,000 each and force them to collaborate, 
even if their philosophies completely clash.
:::


## Poll

<!-- The "URL or I Don't Care" Ultimatum -->

A research group claims to have built an algorithm that predicts early-stage Alzheimers 
with perfect accuracy, but the raw patient data belongs to a private medical corporation 
and legally cannot be open-sourced. What should the scientific community do?

::: {.nonincremental}
A) [Total Rejection:]{.bold-hilit} Ban the paper from academic journals completely; 
unshareable data is fundamentally unshareable science.

B) [The Trust Pass:]{.bold-hilit} Publish it anyway; human health breakthroughs 
are more important than rigid open-science rules.

C) [The Mandatory Summary:]{.bold-hilit} Allow publication only if the authors 
replace the raw data with a heavily aggregated, text-based summary of the data parameters.

D) [The Independent Escrow:]{.bold-hilit} Allow a single independent university professor 
to audit the secret data in a black box on behalf of the public.
:::


## Poll

<!-- The Past vs. The Present -->

Rate your agreement with this statement: 

If an older scientific paper (written before 2000) does not provide a 
working download link to its raw data or script code, it is a waste of 
an undergraduate student's time to read it."

::: {.nonincremental}
1) Strongly Disagree (Brilliant conceptual ideas don't need a download link to be valuable)
2) Disagree
3) Neutral / Undecided
4) Agree
5) Strongly Agree (If I can't replicate or execute it on my laptop, it is a dead end for my education)
:::


## Poll

<!-- The Automated Future -->

Donoho predicts that future data science will be automated by AI bots crawling repos to 
swap parameters robotically. 

\

In exactly **one word**, submit how you feel about a career 
where computers automate the tactical execution of code pipelines.










## Activity

We're going to divide the classroom into groups

. . .

\

:::: {.columns}
::: {.column}
[Team A: The Proprietary Tech Firm]{.mdlit}

You have built a simulation code that calculates the structural stress limits for 
public bridges or medical equipment. You claim it's a corporate secret and refuse 
to share the code lines.
:::

::: {.column .fragment}
[Team B: The Public Safety Board]{.hilit}

You refuse to approve the engineering projects because you cannot verify the 
black-box simulation.
:::
::::


## Activity (cont'd)

- A bridge design built using the firm's secret code has just shown minor structural cracks. 
- The firm claims it's an isolated steel issue, not a software issue.
- They refuse to release their code lines, calling it a multi-million dollar corporate secret.

. . .

Huddle with your immediate neighbors to formulate 3 compelling arguments to defend 
your position.

```{r}
#| echo: false
countdown::countdown(5, bottom = 0)
```

<!--
The Proprietary Tech Firm

1. **[Argument 1: Intellectual Property]**
2. **[Argument 2: Support Overload]**
3. **[Argument 3: Code Modification Risks]**

- Commercial Survival: How do you protect your multi-year financial investment from 
being copied by global corporate rivals]?
- The User Support Trap: If the code goes public, how do you prevent your engineering 
team from being hijacked by strangers demanding free tech support]?
- The Liability Shield: If you open-source the code and someone modifies a parameter, 
causing a crash, who legally owns that failure?


The Public Safety Board:

1. **[Argument 1: Public Accountability]**
2. **[Argument 2: Forensic Transparency]**
3. **[Argument 3: Unverifiable Black Boxes]**

- The Black Box Danger: Why is an un-auditable computational simulation a fundamental 
violation of the scientific method?
- The Life-or-Death Reality: As LeVeque warns, when math moves "outside the ivory tower," 
what happens when code bugs carry real-world body counts?
- The Peer Review Minimum: Why is staring at a final PDF chart or commercial certificate 
completely empty without line-by-line inspection?
-->


## Debate

- Imagine you are a junior researcher. 
- Your script is unoptimized, un-commented, and full of messy hacks.
- But it works: it successfully generates your paper's plots. 
  
. . .

If you publish this "dirty" code:

:::: {.columns}
::: {.column width="45%"}
::: {.poll}
Do you elevate scientific transparency? 
:::
:::

::: {.column width="10%"}
\

&nbsp;&nbsp; OR
:::

::: {.column width="45%"}
::: {.poll}
Do you pollute the scientific record with amateur software engineering? 
:::
:::
::::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

::: {.notes}
If someone emails you for your code 3 years from now, why is that infinitely more 
painful than just dropping your dirty script onto GitHub today?"
:::


## Debate

. . .

Going back to LeVeque's "Theorems Without Proofs" analogy, publication of theorems without
proofs is not acceptable. Yet, we accept computational outputs without the code that 
produce them.

. . .

::: {.poll}
Why does the scientific community tolerate this double standard?
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

::: {.notes}
Is a computer script fundamentally different from a mathematical proof, 
or are we just living in a scientifically immature era of computing?
:::


## Debate

<!-- Protect Your Output vs. Public Provenance -->

::: {.columns}
::: {.column width="50%"}
### 🛡️ The Pragmatist View

It took me three years to write this algorithmic base. If I open-source it now, my 
competitors will use my code to publish five papers before I can even finish my next draft.
:::

::: {.column width="50%"}
### 🌐 The Open View

Publicly archiving your code with a timestamped DOI is the absolute best way to 
claim ownership, establish provenance, and legally force competitors to cite you.
:::
:::

. . .

::: {.poll}
If you are a funding agency, how do you balance a scientist's valid right to profit 
from their multi-year software labor against the public's right to verify the data 
pipeline immediately?
:::

```{r}
#| echo: false
countdown::countdown(4, bottom = 0)
```


## Debate

<!-- The Fear of Maintenance -->

Many researchers hide their scripts because they are terrified that once the code is public, 
their inbox will be flooded with emails from strangers asking for technical troubleshooting, 
custom modifications, or software installation help.

. . .

::: {.poll}
How do you draw a boundary between providing necessary scientific transparency and protecting 
your personal time from being hijacked by outside users demanding free tech support?
:::

```{r}
#| echo: false
countdown::countdown(3, bottom = 0)
```

::: {.notes}
Barnes and LeVeque emphasize that research code is a static archive of a publication, 
not a commercial product that comes with customer support. 
:::


# Quiz-Poll Time

## Google Form

\

**INSERT QR code!!!**

\

**INSERT <url> link to google form!!!**


## Poll

<!--The "Ugly Code" Reality Check -->

You are about to submit a paper. Your code works but is messy, uncommented, and unoptimized. What is your honest action?

::: {.nonincremental}
A) [Publish it immediately:]{.bold-hilit} The dirty code is an essential part of the 
scientific record; transparency wins over aesthetics.

B) [Hide it entirely:]{.bold-hilit} I am too self-conscious about my coding skills; 
I would rather hide the script than face professional judgment.

C) [Delay the paper:]{.bold-hilit} Spend three weeks rewriting it neatly, even if it delays 
my graduation or submission timeline.

D) [The "Upon Request" Lie:]{.bold-hilit} State that the code is "available upon request" 
and hope no one ever actually emails me.
:::


## Poll

<!-- The Computational Double Standard -->

Why does the scientific community tolerate papers that publish computational figures but hide the source code scripts?

::: {.nonincremental}
A) [Cultural Lag:]{.bold-hilit} Computing is relatively new to academia; we are simply living in a scientifically immature era of data science.

B) [Complexity:]{.bold-hilit} A computer script is infinitely more complex than a static math proof, making routine auditing impractical.

C) [Status Bias:]{.bold-hilit} We evaluate science based on the prestige and reputation of the author's university rather than their code.

D) [Journal Laziness:]{.bold-hilit} Editors and peer reviewers simply refuse to force authors to change their habits because it adds extra work.
:::


## Poll

<!-- The "Scooping" Threat Matrix -->

Rate your level of agreement with this statement: 

"A researcher has a right to keep their novel code hidden for up to two years after publication to ensure competitors don't scoop their future research ideas."


::: {.nonincremental}
1) Strongly Disagree (Public tax dollars funded it; code belongs to the world immediately)
2) Disagree
3) Neutral / Undecided
4) Agree
5) Strongly Agree (Career survival metrics must outweigh open-science ideals)
:::


## Poll

What is your biggest psychological fear regarding open-sourcing your personal computational 
research project?

\

In exactly **one word**, submit your primary worry or mental roadblock regarding making 
your code visible to the public.


# Final Words



## Key Takeaways

- Open science isn't just an academic debate—it has high-stakes civic consequences.

- Hidden code can carry direct life-or-death consequences in engineering and public policy.

- Perfectionism and Absolute Secrecy both break scientific progress. 


## Key Takeaways

In the real world, engineering projects navigate this standoff through compromised 
structures:

- **The Production Sandbox:** Tech firms can deposit source code into a 
    *Software Escrow*—allowing a trusted, independent third-party academic panel to 
    run line-by-line audits without exposing trade secrets to public competitors.

- **The Barnes Reality Check:** Code does not need to be a beautifully polished 
    product to be verified; it just needs to expose the raw mathematical parameters 
    governing the bridge's stress metrics.

- **The Baseline Rule:** Un-auditable code is un-verifiable science.


## The Selfish Benefits of Open Code

- [Your Most Important Collaborator:]{.bold-hilit} Your "past self" is your most frequent 
collaborator, and they notoriously forget what their own code did after 6 months.

- [The "Push-Button" Advantage:]{.bold-hilit} Forcing yourself to build a script that 
replicates your whole project with the push of a button pays off massively down the road.

- [Frictionless Revisions:]{.bold-hilit} When a journal reviewer inevitably demands you 
replot your data or shift an axis, an automated script saves you weeks of manual 
reconstruction labor.

- [The Institutional Shield:]{.bold-hilit} If your results are ever publicly questioned 
or audited by outsiders, a public, timestamped git repository provides immediate, 
undeniable proof of your data's provenance.