---
title: "Reading and Discussion"
format: stat159slides-revealjs
chalkboard: true
---

# Reproducible Research Standard (RSS) Framework

## Reading

::: {.nonincremental}
- Title: **The Legal Framework for Reproducible Scientific Research**
- Author: Victoria Stodden
- Journal: Computing in Science & Engineering
- Year: 2009
- <https://www.stodden.net/papers/Legal-STODDEN2009.pdf>
:::


## About

Victoria Stodden proposes the Reproducible Research Standard (RRS). 

It aligns intellectual property law with open scientific norms.


# Reading's Key Points

## The Legal Conflict

- Default copyright applies automatically to digital work.
- Copyright laws grant exclusive monopolies to creators.
- This legal barrier blocks computational reproducibility.
- Scientific progress requires continuous modification and reuse.
- Standard copyright structures run counter to scientific ethos.


## The Research Compendium

- Academic papers are merely advertising for scholarship.
- True scholarship includes the complete development environment.
- A full research compendium requires five components:
    + Manuscript source files.
    + Data and collection documentation.
    + Experimental source code scripts.
    + Final figures and processing logs.
    + Auxiliary web presentation materials.


## Copyright and Data

- Raw objective facts cannot be copyrighted.
- Applying standard copyright to facts is meaningless.
- Data curation requires immense time and labor.
- Original dataset selection and arrangement can hold copyright.


## The Share Alike Problem

- Copyleft licenses mandate identical downstream licensing.
- Share Alike clauses severely restrict open scientific reuse.
- They block commercial industry from using academic research.
- They complicate proper author attribution in derivative works.
- Science needs credit without restrictive licensing bounds.


## The Reproducible Research Standard (RRS) Solution

- The RRS acts as an umbrella licensing algorithm.
- It separates a compendium into distinct asset categories.
- It applies specific open frameworks to each asset:
    + Modified BSD for code scripts.
    + Creative Commons BY for media and narrative text.
    + Science Commons Database Protocol for public domain data.
- It mandates viral attribution for all components.
- It protects scientists from uncredited intellectual theft.


# Share Alike Issues

## Why Share Alike Blocks Industry

- The Forced Open-Source Trap: Companies must open-source their entire product 
if they use your code.

- Financial Risk: Corporate legal teams strictly ban engineers from using 
Share Alike assets.

- Zero Adoption: Your research gets ignored by industry developers due to 
legal liability.

- Kills Tech Transfer: Universities cannot license the technology to 
commercial startups.


## Why Share Alike Hurts Science

- License Incompatibility: You cannot mix two datasets if they have different 
Share Alike terms.

- Attribution Stacking: Combining multiple works makes it impossible to track 
clean credit.

- The Instrument Barrier: Academic code often must run inside proprietary software 
like MATLAB.

- Scientific Norms: Science prioritizes the spread of truth over forcing specific 
licenses.


## Stodden's Alternative: "Viral Attribution" 

- Instead of forcing others to share their entire project, Stodden advocates 
for mandatory credit. 

- If a company uses your academic code, they can keep their product secret, 
but they must legally cite your paper and give you credit. 

- This satisfies the academic reward system (citations) without killing economic 
innovation.


# Attribution Clause

## Attribution Clause

According to Stodden's framework, a strong attribution clause is not only 
enough---it is actually the ideal mechanism for scientific compliance.

Here is why Stodden argues that attribution alone is the correct legal tool 
for open science.


## It Aligns with Existing Scientific Norms

The Currency of Science: In academia, the primary reward system is credit, 
impact, and citation. 

Scientists do not usually lose sleep over whether someone made money off 
their idea; they care if they were plagiarized or left uncredited.

Legal Enforcement of Citations: Traditional academic citation is just a 
social norm. By using an attribution-only license (like CC BY or modified BSD), 
you turn a polite scientific etiquette into a legal requirement. 

If someone uses your compendium without citing you, they are committing 
copyright infringement.


## "Viral Attribution" Without the "Share Alike" Damage

- Stodden introduces the concept of viral attribution.
- This means the license requirement to give credit propagates forward 
to any downstream work that builds on your research.
- Unlike Share Alike, it only forces the downstream user to credit your 
specific component. 
- It does not force them to change the license of their own unique, 
proprietary add-ons or comingled work. 
- Unlike Share Alike, it only forces the downstream user to credit your 
specific component. It does not force them to change the license of their 
own unique, proprietary add-ons or comingled work.


# Discussion Time

## Scenario A

Your lab neighbor downloads a clean, un-licensed Python script from a 
public GitHub repo, fixes a minor bug, and uses it to generate the main 
figure for a high-impact paper.

::: {.poll}
Did your classmate commit copyright infringement? If the paper is published, 
is the science legally reproducible?
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

<!--
Answer: Yes, it is technical infringement. Remind the class that under the 
Berne Convention, "Public to see" on GitHub does not mean 
"Public to copy and modify" locally.
<--


## Scenario B

You wrote a beautiful data-scraping algorithm in your university dorm room 
using your personal laptop, but you did it while receiving a university 
stipend or working as a graduate teaching assistant.

::: {.poll}
Who actually holds the automatic default copyright to that code? 
You, or the university Board of Trustees?

If you want to open-source it under an MIT license, do you legally have 
the right to sign that document?
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

<!--
Highlight the reality of Institutional Employment Agreements. 
Most universities claim ownership over intellectual property created 
by employees (including RAs/TAs) using university resources, which makes 
open-source licensing an institutional negotiation, not just a personal choice.
-->


## Scenario C

A student wants to release their climate change simulation dataset but 
insists on adding a "Non-Commercial" restriction because they don't want 
ExxonMobil or a private oil firm making money off their hard academic work.

::: {.poll}
Who else gets accidentally blocked by an "NC" clause?
:::

```{r}
#| echo: false
countdown::countdown(2, bottom = 0)
```

<!--
Point out that private universities (Stanford, Harvard), non-profit medical 
centers, or publicly funded labs with commercial partners are legally blocked 
from reproducing this student's science due to that well-intentioned clause.
-->


# Class Activities

## Activity 1

Consider a real-world scientific repository containing 4 distinct files:

1) `model_simulation.py` (The logic script)
2) `patient_clinical_trials.csv` (Raw data matrix)
3) `manuscript_draft.qmd` (The text narrative)
4) `results_plot.png` (The visual graph)

. . . 

\

The Challenge: Fix a legally broken, computationally corrupted research repository.

<!--
Pass out cards (or use a digital poll like Mentimeter) with three tools: 

- [MIT/BSD], 
- [CC-BY 4.0],
- [CC0/Public Domain].

The Race: Teams must correctly match exactly one tool to each file according 
to Stodden's RRS framework as fast as possible.
-->


## Activity 2

Consider another project with the following components:

a) Repo has no `LICENSE` file
b) Raw data has a CC BY-NC-ND license
c) The main script relies on a secret, proprietary tool

. . .

With the given cards, find the one that solves each issue.

<!--
Solutions: 

a) Use (MIT or BSD) to legally copy the code to their laptop.
b) Change license to CC0.
c) Swap tool for an open-source equivalent.
-->


# Quiz-Poll Time

## Google Form

\

**INSERT QR code!!!**

\

**INSERT <url> link to google form!!!**


## Poll

<!-- The Proprietary Ultimatum -->

If a scientific pipeline relies on an encrypted corporate file, commercial software, 
or sensitive dataset that legally cannot be shared, what should happen?

A) [Total Rejection:]{.bold-hilit} The paper should be banned from scientific journals; 
unshareable code is fundamentally unshareable science.

B) [Conditional Acceptance:]{.bold-hilit} Allow publication as long as the authors 
explicitly write out a text description of what the hidden code does.

C) [Partial Openness:]{.bold-hilit} Mandate that they open-source *only* the specific 
custom lines they wrote themselves, even if the base system remains a black box.

D) [Full Pass:]{.bold-hilit} Trust the authors; proprietary dependencies are a normal 
reality of working with industry partners.
